\batcave: Calling somatic mutations with a tumor- and site-specific prior

### 2019-12-19 ResponseToReviews.tex

We appreciate the opportunity to revise our manuscript for publication in \emph{NAR Genomics and Bioinformatics}, and we thank you and the reviewers for the quality of the reviews. Below we respond in italics to the reviewers' concerns.

\noindent Sincerely,\\
\includegraphics{signature}\\
Ryan Gutenkunst, PhD\\
Associate Professor and Associate Department Head\\
Department of Molecular and Cellular Biology\\
University of Arizona

\section*{Associate Editor}

Stressing on some of the points raised by the reviewers, I would suggest you include more details about the assumption and limitations of BATCAVE. In particular, the possibility of changes of mutational signatures over time can be quite striking in some cases, especially if the original tumour is driven by very specific mutagens like UV light or tobacco smoke. Ideally, the software should try to correct for these cases or at least warn the user about a possible shift in signatures/contexts.
\begin{response}
As we noted in the Discussion, BATCAVE does assume that the mutation profile is similar between mutations that are high-confidence (tend to be earlier) and low-confidence (tend to be later).
Modeling potential changes in the profile is beyond the scope of the current manuscript, but warning the user if such a shift exists is a great idea.
We have thus implemented a statistical test as to whether the mutation profile from the high-confidence mutations differs significantly from the low-confidence mutations.
If this test detects a difference, BATCAVE warns the user and outputs both profiles, so the user can assess the degree of difference and follow-up on any potential biological implications.

When we reanalyzed our test data using this new approach, we did detect a profile shift in the AML data.
Note that even with this shift, BATCAVE still performs better than MuTect on the validated mutations (although both perform extremely well).
\RNG{The biology of the shift is...}
\end{response}

@@

Mannakee and Gutenkunst present BATCAVE, an algorithm that first estimates the tumor’s mutation profile and mutation rate using high-confidence variants and then uses them as a prior when calling other variants based on mutational signatures. The concept behind this work is interesting. However, the difference in accuracy between MuTect and BATCAVE is relatively small except for the case when one has particularly high depth sequencing (500x), limiting the utility of the method.

Still, the 500x whole exome case does indicate that BATCAVE is better at identifying low frequency variants and thus can better reveal intratumoral heterogeneity, an important area of inquiry in cancer. Interestingly, the AUROC values for both MuTect and BATCAVE have room for improvement for the 500x whole exome case (Table 1). I suggest the authors further investigate how to improve the AUROC values. For example, in equation 4, it is unclear why the authors assume $\nu$ be uniformly distributed. Although this is also an assumption of MuTect, the BATCAVE approach inherently distinguishes mutations by allele frequency – since log-likelihood is closely associated to AF.  Perhaps a model in which $\nu$ follows a decreasing distribution vs. AF (e.g. in concordance with a model of a growing tumor) would yield better AUROC values.  
\begin{response}
A uniform prior over $\nu$ is appropriate when the goal is to make the best possible call at each potential variant site.
Conceptually, before examining the read data we have no information about the allele frequency at any individual site.
Practically, a nonuniform prior would favor low-frequency and thus typically low-evidence sites, increasing the false positive rate.
If our goal were to best estimate the distribution of variant allele frequencies in the tumor, then a nonuniform $\nu$ would indeed be important.

We discuss this issue in a new paragraph in Discussion.
\end{response}

In Rubanova et al ( https://www.biorxiv.org/content/10.1101/260471v4.full), it is argued that there are changes in mutational signatures over time in cancers. While Mannakee and Gutenkunst mention this work, they should further comment on whether such changes in mutational signatures over time are small enough to be explained by the BATCAVER concept of re-evaluating mutations based on p-value.
\begin{response}
We cannot comment generally on the magnitude of mutation profile changes, because this is an developing area of research.
But to protect users, we have implemented a new statistical test to detect changes in mutation profiles between high- and low-confidence mutations.
See our response to the Associate Editor for more details.
\end{response}

### 2019-12-19 NAR_DRAFT.tex
Add draft responses to profile change comments.

\subsection{Implementation}
We have implemented the \batcave algorithm as an R package \texttt{batcaver}.
The package leverages the Bioconductor packages \textrm{BSgenome} \cite{Pages2019}, \textrm{GenomicAlignments} \cite{Lawrence2013}, \textrm{VariantAnnotation} \cite{Obenchain2014}, and \textrm{SomaticSignatures} \cite{Gehring2015} for fast and memory-efficient variant annotation and genomic context identification.
Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.

@@

Griffith et al.\ sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>$360X and used targeted sequencing to validate nearly 200,000 mutations~\cite{Griffith2015}.
We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which is consistent with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
For both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).

Shi et al.\ performed multi-region whole exome sequencing on six individual breast tumors to a mean target sequencing depth of 160X and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
%We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated an average per-base mutation rate for these tumor regions of $4\cdot10^{-8}$, which is consistent with observed mutation rates for breast cancers \cite{Alexandrov2019} and with the low number of validated somatic mutations.
For the validated variants, MuTect and \batcave yielded almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})

@@

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior and mutation rate, which are then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\cite{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \cite{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
An  implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, and although in general those differences are relatively small, they can be significant in some cases.
\bkmcomment{We provide the user with the ability to output a file with the contents of the estimated mutation profile as each high-confidence variant is evaluated.
In cases where the user suspects a shift in mutational process, we encourage using this file to verify that the final mutation profile is representative of the processes active at low frequency.}
A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.

@@

% \section{CONCLUSION}
% \bkmcomment{I looked at a bunch of NAR methods papers, and none include a conclusions section, so I moved it to the last paragraph of discussion.}

@@

>>>>

\subsection{Implementation}
We have implemented the \batcave algorithm as an R package called \texttt{batcaver}.
The package leverages the Bioconductor packages \textrm{BSgenome} \cite{Pages2019}, \textrm{GenomicAlignments} \cite{Lawrence2013}, \textrm{VariantAnnotation} \cite{Obenchain2014}, and \textrm{SomaticSignatures} \cite{Gehring2015} for fast and memory-efficient variant annotation and genomic context identification.
Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.

\rngcomment{
An assumption of the \batcave algorithm is that the mutation profile of low-confidence mutations is similar to that of the high-confidence mutations from which the prior is inferred.
To test this assumption, \texttt{batcaver} performs a statistical hypothesis test comparing these two mutation profiles \cite{Plunkett2019}.
If the two profiles differ at a significance threshold of $\alpha=XXX$, \texttt{batcaver} issues a warning and outputs the profiles.
%The user can then assess them and follow-up on any biologically interesting differences.
}

@@

Griffith et al.\ sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>$360X and used targeted sequencing to validate nearly 200,000 mutations~\cite{Griffith2015}.
We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which is consistent with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
\rngcomment{
For this tumor, the mutation profiles of high- and low-confidence mutations differ ($p \sim 10^{-12}$).
The low-confidence mutation profile contains additional spikes of mutations corresponding to XXX (Fig.~\ref{fig:AML_profiles}).
}
Nevertheless, for both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).

Shi et al.\ performed multi-region whole exome sequencing on six individual breast tumors to a mean target sequencing depth of 160X and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
\rngcomment{Brian: Add purity estimate, if in paper.}
%We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated an average per-base mutation rate for these tumor regions of $4\cdot10^{-8}$, which is consistent with observed mutation rates for breast cancers \cite{Alexandrov2019} and with the low number of validated somatic mutations.
For the validated variants, MuTect and \batcave yielded almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})

@@

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior and mutation rate, which are then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\cite{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \cite{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.

An  implicit assumption of \batcave is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, and although in general those differences are relatively small, they can be significant in some cases.
\rngcomment{Our \texttt{batcaver} software warns the user if the high- and low-confidence mutational profiles are statistically different.
It also outputs those profiles, so the user can assess whether the difference will substantially affect their results.} 
A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.

@@

@@

\begin{figure} [H]
\centering
  %\includegraphics{XXX}
  \caption{Mutation profiles of high-confidence and low-confidence mutations in the AML data.}
\label{fig:AML_profiles}
\end{figure}
